---
title: 'Programming assignment, R: Simulating Pathways, Mutual Exclusivity, et al. in Cancer Progression Models'
author: 'Raquel Blanco Martínez-Illescas^[Universidad Autónoma de Madrid, Bionformatics and Computational Biology Master], Daniel Peñas Utrilla$^*$, Henry Secaira Morocho$^{*}$'
date: "`r Sys.Date()`"
params:
  doc_title: "Programming_R_assignment"

output:
  pdf_document:
    toc: yes
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    toc_depth: 3

bibliography: Bibliografia_R.bib
link-citations: yes
linkcolor: blue
csl: vancouver.csl
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# \textcolor{blue}{Introduction} {#one}

## \textcolor{blue}{Cancer Progression Models} {#one_one}

Cancer is an heterogeneous disease caused by the continuous accumulation of different somatic mutations during lifetime of an individual [@raphael_simultaneous_2015, @Schill2020]. This somatic mutations include Single Nucleotide Variants (SNV), Copy Number Variations (CPV) or insertions and deletions (INDELS). However, not all mutations contribute to malignant initiation or cancer progression. Therefore, identifying mutation leading to cancer progression becomes key to understand cancer development and treatment [@Pon2015]. Somatic mutations that affect the cells are classified in two main groups: *passenger* mutations and *driver* mutations. Passenger mutations are silent mutations which doesn't lead to any pathogenic situation (these are the main alterations affecting cells). On the other hand, driver mutations are responsible for cancer onset. Those mutations provide the cells with morphological and metabolic alterations that ultimately lead to a selective growth advantage over wild type cells from which they derived [@Nowell1976]. 

Driver mutations can affect both oncogenes and tumor suppressor genes, but conversely. Oncogenes are altered proto-oncogenes leading to an uncontrolled growth rate, division and survival of cells. Thus, specific mutations promoting over-activation of these genes will yield to cancer onset. On the other hand, tumor suppressor genes prevent unrestrained cellular growth and promote DNA repair by protein recruitment. Furthermore, lower or abnormal expression of these genes threat DNA repair mechanisms and growth homeostasis [@Lee2010]. 

Cancer progression is caused by the accumulation of various mutations not just simply one mutation as occurs in monogenic diseases. Moreover, not all possible mutations order seem to be equally responsible for cancer progression. Indeed, some mutations require some previous alteration to happen (dependencies between genes). Therefore, it is necessary to know which are the constraints and restrictions that lead to cancer development. Models explaining those dependencies are called **Cancer Progression Models (CPMs)** [@Diaz-Uriarte2015]. CPMs are depicted as Directed Acyclic Graphs (DAG). In this graph, nodes represent genes and arrows dependencies between them. The path refers to the order of mutations to cancer progression [@Diaz-Uriarte2018].

Although cancer progression is a dynamic process, tumor data is usually obtained as **cross-sectional** samples. It is a combination of snapshots taken from different tumor progression at different stages [@Schill2020]. Assuming that these observations reflect the same stochastic process, they can be used to infer restrictions between tumor events. This order of effects can be report as a direct graph. Knowing this constraints between genes, it allows to define a therapeutic target to avoid cancer progression [@Diaz-Uriarte2018]. However, temporal progression of any cancer is difficult to infer by cross-sectional data. The utopian dataset to infer cancer progression is a *longitudinal dataset* consisting of same individual tumor samples from different time points during cancer development. Nevertheless, it is almost impossible without damaging individuals [@raphael_simultaneous_2015].


## \textcolor{blue}{Evolutionary Models} {#one_two}
(Just and idea) Mention the evolutionary models and how OncoSimulR allows deviations from monotonicity with genotype to fitness mapping

## \textcolor{blue}{Order of effects} {#one_three}
(Just and idea) Describe order of effects

## \textcolor{blue}{Epistatic interactions} {#one_four}

Mutual exclusivity is a common phenomena in cancer progression. It is defined as the situation where two events (mutations) occur less frequently than expected by chance [@Schill2020]. Two genes are mutually exclude if the presence of one of them avoids the presence of the other. In nature, there are two mechanisms that can lead to this phenomena: 

- Synthetic lethality, where carrying both mutations is detrimental for the viability of the cell.

- Null effect, where whichever mutation occurs first involves most of the selective advantage and decrease the selective pressure to occur for the others.

### \textcolor{blue}{Synthtetic Viability} {#one_four_one}
(Just and idea) Describe sign synthetic viability

### \textcolor{blue}{Mutual exclusivity} {#one_four_two}
(Just and idea) Here mention: Mutual exclusivity for Null effects and synthetic lethality (reciprocal sign epistasis)

## \textcolor{blue}{Frequency dependent fitness} {#one_five}
(Just and idea) Describe frequency dependent fitness

<!-- Now the models of each one! -->

# \textcolor{blue}{pathTiMEx, a generative probabilistic graphical model of cancer progression} {#two_one}

In @Cristea2017, the authors introduce a generative probabilistic graphical model of cancer progression called *pathTiMEx*. It is both, a waiting time model for independent mutually exclusive pathways, and a waiting time model for cancer progression among single genes. 

This generative model allows to generate a cancer progression model including mutual exclusivity between groups and progression among pathways. In this approach, authors think in both, genes and modules effects (set of genes). 

The colorectal cancer model depicted in Figure 3.A [@Cristea2017] is used as an example of model to map. The colorectal cancer dataset used to built that model is obtained from @Wood2007. The poset restrictions proposed can be coded using the **OncoSimulR package** [@Diaz-Uriarte2017], concretely, the `allFitnessEffects` function. It creates mutations effects given specification of restrictions, epistasis or order effects. In this case, restrictions are used to construct the graph. 

Some parameters are mandatory when `allFitnessEffects` function is used. It is the case of the *restriction table*. It specifies the dependencies between genes or modules (genes/modules parents and genes/module children). Parameter `s` and `sh` refers to a numeric vector with the fitness effect that applies if the relationship is or is not satisfied, respectively. Authors don't specify its value since they are not interested in fitness. To justify the values given, we will use the waiting time rate parameter \(\lambda\) defined in the model. Early events in cancer progression will have greater \(\lambda\) values while late events will have a lower one (values for all genes or modules are showed in \autoref{table1}). Thus, genes/modules with higher \(\lambda\) will receive a higher fitness value (`s`). On the other hand, `sh` is given a constant value for all possible situations.

Table: Waiting time rate parameter (\(\lambda\)) for each gene/module \label{table1}

| Gene/module | Waiting time rate parameter (\(\lambda\)) |
| :-----------: | :-------------------------------------: |
|      APC      |                  9.5                    |
|               |                                         |
|     KRAS      |                  2.89                   |
|               |                                         |
|  TP53, EVC2   |                  1.92                   |
|               |                                         |
| PIK3CA, EPHA3 |                  0.17                   |
|               |                                         |
| FBXW7, TCF7L2 |                  0.08                   |

Although authors don't specify the sort of dependency, in this poset, a semimonotonic dependency is defined between modules B and C with E (`SM`), while a monotonic dependency is defined for the others (`MN`). Model will be represented as an **Diaciclic Direct Graph (DAG)** where arrows conecting genes or modules indicate direct dependencies or constraints between them [@Diaz-Uriarte2018].

```{r fig1, fig.align="center", fig.cap= "DAG from colorecta cancer"}
## First, it is necessary to load OncoSimulR and igraph package
library(OncoSimulR)


## Restriction table (extended version of the poset)
colcancer <- data.frame(
                 parent = c(rep("Root",3), "A", "B", "C"), # Parent nodes
                 child = c("A", "B", "D", "C", "E", "E"), ## Child nodes
                 s = c(rep(0.5, 3), rep(0.05, 3)), 
                 
                 sh = -0.5, 
                 
                 typeDep = c(rep("MN", 4), rep("SM", 2)) ## Type of dependency 
                  )

## Fitness specification of the poset
colcancer_efec <- allFitnessEffects(
                  colcancer, # Poset
                  
                  geneToModule = c( ## Specification of the modules
                               "Root" = "Root", 
                               "A" = "APC",
                               "B" = "TP53, EVC2",
                               "C" = "KRAS",
                               "D" = "PI3KCA, EPHA",
                               "E" = "FBXW7, TCF7L2"),
                  
                  drvNames = c( ## Specification of drivers
                               "APC", "TP53", "EVC2", "KRAS",
                               "PI3KCA", "EPHA", "FBXW7", "TCF7L2")
                  )



## DAG representation
plot(colcancer_efec, expandModules = TRUE, autofit = TRUE, lwdf = 2)

```

Model proposed by @Cristea2017 indicates that, from a wild type genotype (depicted as "Root" in \autoref{fig:fig1}) it is possible to follow three different paths. The wild type genotype can suffer a mutation in the `APC` gene or in the `TP53`, `EVC2` or `PI3KCA`, EPHA modules. These mutations occur independently, without the need for previous mutations to occur.

In the case of the `PI3KCA`, `EPHA` module, it is not essential for other mutated genes or modules to appear. On the other hand, the mutation in the APC gene is necessary for a mutation to occur in the `KRAS` gene. The `APC` gene is the parent node of the `KRAS` gene (**monotonicity dependency**). Also, the `KRAS` gene would be necessary for the mutation to appear in the `FBXW7`, `TCFL2` module. However, the relationship of the `KRAS` gene to the `FBXW7`, `TCF7L2` module has been defined as a **semitonicity dependency**. This implies that there is no need for a mutation in the `TP53`, `EVC2` module in order for it to mutate. Similarly, a mutation in the `TP53`, `EVC2` module would be sufficient for a mutation in the `FBXW7`, `TCF7L2` module to occur (this semitonicity dependency is dipected in blue color in \autoref{fig:fig1}).


In a DAG, only the genotypes that fulfill the restrictions defined by the arrows connecting the genes/modules can exist. Moreover, restrictions set in DAG do not contain any information about fitness of each individual genotypes, it is just a pathway to follow to cancer progression. On the other hand, fitness landscapes (or genotype-fitness maps) show the fitness associated to each genotype allowing to know the impact of a specific mutation. Furthermore, the restrictions reflected in the DAG just show sign epistasis between genes/modules. Epistasis is an effect where the phenotypic consequences of a certain mutation depend on the genetic background in wich it take place. 

A specific type of epistasis is called *sign epistasis* and it refers to the case where a mutation yields to a increase (beneficial) or decrease (deleterious) of the fitness depending on the genotypic background of the cell. It has a different sign depending on the other genes mutated in the clone cell. Although this kind of epistasis can be depicted using a DAG, it can not show reciprocal sign epistasis, a specific situation where two individual mutations increase the fitness, although combine reduce it (synthetic lethality). 

Hence, to visualize the relationships between genotypes and effects in fitness, the fitness landscape using the restrictions specified in \autoref{fig:fig1} is generated. For that aim, the `evalAllGenotypes` function is used. It returns a table with all the genotypes from the fitnessEffects description indicated as well as the genotype associated to it. The table obtained can be used as an object to plot a fitness landscape of the \autoref{fig:fig1}.

```{r fig2, fig.align="center", fig.cap="Fitness landscape from colorectal cancer"}
colcancer_efec_FL <- evalAllGenotypes(colcancer_efec, max = 110000)
## Output is not shown due to size of the table.

## Plot of fitness landscape
plotFitnessLandscape(colcancer_efec_FL)
```

Fitness landscape obtained is displayed in \autoref{fig:fig2}. It shows a quite busy fitness landscape due to the huge amount of possible genotypes combinations, each one with a different fitness value. However, it reflects genotype acquisition in terms of survival and adaptation for the cell, allowing to set what is the fitness associated to a clone that gets a certain mutations or a certain number of them. 



## \textcolor{blue}{Simplified cancer progression model} {#two_two} 

In order to properly visualize a fitness landscape, a simplified version of the model coded in \autoref{two_one} is constructed. This model doesn't use modules, just individual genes. This approach will lead to clear fitness landscape and to properly identify processes that may occur.

Authors [@Cristea2017] claim that there is a phenomena of mutual exclusivity between certain genes of specific pathways. Mutual exclusivity means that the presence of one gene in a specific pathway will be enough to fitness contribution, since mutation in the other genes of the same pathway are negative selected and therefore, the presence of the other gene in the same module can be discarded, since they will not mutate.

Therefore, the same model as in the previous case will be coded, but without specifying modules. Each module will be consider as an specific gene.

```{r poset}

## Fitness specification of the simplified poset
Scolcancer <- allFitnessEffects(colcancer)
```

```{r fig3, fig.align="center", fig.cap="DAG from a simplified model of colorecta cancer"}
plot(Scolcancer)
```

```{r table2, fig.align= "center"}
## Obtain all genotypes from the fitnessEffect
(Scolcancer_ge <- evalAllGenotypes(Scolcancer))
```

```{r fig4, fig.align="center", fig.cap="Fitness landscape from a simplified model of colorectal cancer"}
## Plot the fitness landscape.
plotFitnessLandscape(Scolcancer_ge,
                     use_ggrepel = TRUE)

```

DAG graph and fitness landscape of this simplified model is depicted in \autoref{fig:fig3} and \autoref{fig:fig4}, respectively. DAG showed in \autoref{fig:fig3} is the same as the DAG depicted in \autoref{fig:fig1} but without expanding modules. In this case, there is not an improvement in legibility or clarity. However, if we compare the fitness landscape obtained with the simplification (see \autoref{fig:fig4}) with the previous fitness landscape (see \autoref{fig:fig2}), there are a clear difference in clarity. In this new fitness landscape is possible to visualize the fitness given to each genotype and therefore, give an evolutionary sight to the model. 

## \textcolor{blue}{Simulating data from a simplified model} {#two_three}

Restrictions set in DAG were used as a guide line to built the fitness landscape (see \autoref{fig:fig4}). This fitness landscape shows each possible genotype as well as its fitness. This landscape can be used to simulate fitness evolution in cancer progression. `OncoSimulIndiv` function is used to simulate colorectal tumor progression. This function simulates a single evolutionary path. It is necessary to include the poset with the order restrictions defined for the simplified model (see \autoref{two_two}). McFarland model (continuous-time, logistic-like, and death rate depends on population size) is used for simulation of cancer progression, since it leads to a better performance [@Diaz-Uriarte2015]. Initial population size is set at 400. Only one mutation rate is used: 1e-4. Final time is set to 220 to visualize clones' evolution. Furthermore, keepPhylog parameter is set true to plot the parent-child relationships occurring in the simulation as well as its frequency (`plotClonePhylog` function).

```{r fig5, fig.align= "center", fig.cap= "Simulation of cancer progression using the fitness landscape of the simplied model (stacked plot)"}

set.seed(257) ## Fix the seed for reproducibility

Simul <- oncoSimulIndiv(Scolcancer, ## A fitnessEffects object
                      model = "McFL", ## Model used
                      mu = 1e-4, ## Mutation rate
                      sampleEvery = 0.02, ## How often the whole population is sampled
                      keepEvery = 1,
                      initSize = 400, ## Initial population size
                      finalTime = 220,
                      keepPhylog = TRUE, ## Allow to see parent-child relationships
                      onlyCancer = FALSE
)


## Plot of simulation
plot(Simul, ## OncoSimulIndv model
     show = "genotypes",
     type = "stacked"
     
)
```


```{r fig6, fig.align= "center", fig.cap= "Simulation of cancer progression using the fitness landscape of the simplied model (line plot)"}
## Plot of simulation
plot(Simul, ## OncoSimulIndv model
     show = "genotypes",
     type = "line"
)
```

```{r fig7, fig.align="center", fig.cap="Parent-child relationship derived from simulation"}
## Parent-child relationship derived from simulation
plotClonePhylog(Simul,
                N = 0, ## Specify clones that exist
                keepEvents = TRUE ## Arrows showing how many times each clones appeared
                
                
)
```

A stacked and line plot of the simulation is depicted in \autoref{fig:fig5} and \autoref{fig:fig6}, respectively. Both plots show the genotype acquisition by time and the number of clones carrying that genotype in the cell culture. Different cell population converge in different time moments, each carrying a diferent genotype and therefore, a different fitness. 

In \autoref{fig:fig5} wild type genotype ("WT") progressively despairs while clones carrying a mutation in gene "A" arrive to culture. However, they are substituted by a new clone that also carries a mutation in "B". Then, this clone suffers different mutations resulting in the coexistence of different genotypes in the cell culture, each one with a different fitness. Finally, genotype carrying all genotypes is stabilized in the culture. As it fitness is the greatest among all the fitness of all genotypes. It is obvious that it is more selective favored and it will overcome the concurrence with the other genotypes.

On the other hand, \autoref{fig:fig6} shows the same information but it is possible to observe all genotypes generated in the simulation, even those that survive for little time. In addition to the genotypes seen in \autoref{fig:fig5}, some other genotypes appear in the cell culture, but due to selective pressure, they are not able to survive. 

Hence, genotypes observed in the simulation not only follow the restrictions set in DAG of \autoref{two_one}. It is just a generative model where dependencies between genes are defined, but they may not occur in real life in that order. Moreover, this model is constructed using cross-sectional data, tumor snapshots in a specific time of cancer progression, it is not a temporal dataset from the onset to the end of the cancer progression.

\autoref{fig:fig7} shows the genotype evolution in the simulation. Arrows' width represent frequency of clone creation. Widther arrows indicate a higher frequency of change from the parent genotype to the child genotype.

## \textcolor{blue}{Order effects} {#two_four}

To explore order effects in cancer progression, a simple model derived from the restriction model inferred by @Cristea2017 is created. 

This simplified model just contains 3 genes: `APC`, `TP53` and `KRAS`, genes considered as `superdrivers` [@Gerstung2011], meaning that are the main responsible for cancer progression since they provide a higher fitness gain than the other genes in the model. This conclusion is obtained from its article where they used the same colorectal cancer dataset as @Cristea2017. Thus, it can be extrapolated to our case.

The relationships between those genes was previously depicted in \autoref{two_one}. In this case, we will set `APC`  as the parent of `KRAS`. Both, `APC` and `TP53` have as parent `Root`. Based on the waiting time rate parameter \(\lambda\), the fitness values of each possible order is given (see \autoref{table1}). 

\(\lambda\) is higher for `APC`, which means that it seems to appear before in the cancer progression. \(\lambda\) for `KRAS` is the lower between the three, meaning that it mutates the last. `TP53` mutation occurs between `APC` and `KRAS`. Order effects are defined following this criteria: clones suffering mutations in the previous order are favored with a higher fitness. Other possible paths of cancer progression are slightly less naturally selected (assumption based on @Cristea2017). Order effect is visualize using `evalAllGenotypes` function.

```{r}

cc <- data.frame(parent = c(rep("Root", 2), "A"),
                 child = c("A", "C", "B"),
                 typeDep = "MN")

cc_order <- allFitnessEffects(
  orderEffects = c("A > B > C" = 0.5, "B > A > C" = 0.2,
                   "B > C > A" = 0.1,
                   "B > C" = 0.2,
                   "C > B" = 0.1,
                   "B > A" = 0.1,
                   "A > B" = 0.3),

  geneToModule =
    c("A" = "APC",
      "B" = "KRAS",
      "C" = "TP53") )

(cc_order_geno <- evalAllGenotypes(cc_order, order = TRUE))

```

We obtain a table with the different possible genotypes as well as the order of appearance. However, this approach doesn't allow to generate neither a DAG neither a fitness landscape. Thus, is not possible to visualize the evolution of the genotypes with time. 

```{r, error=TRUE}
#DAG
plot(cc_order)

# Fitness landscape
plotFitnessLandscape(cc_order_geno)

```

Assuming a model where there is not an order effect, a mutation in gene "B" followed by a mutation in gene "A" will reach the same fitness as if the mutation in gene "A" occurs first. However, in the model just generated, the order of the mutation impacts the final fitness reached by the tumoral clone. Since, the previous alteration of some genes before can lead to an evolutionary advantage. 

In a non order effect model, the final fitness value is the same for all the clones, while this is unlikely to happen in real life. Clones carrying a certain mutation from the beginning would survive easily than those reaching the same genotype in a different order.

This is one limitation of `OncoSimulR package`, it doesn't allow to visualize those scenarios (yet). 

## \textcolor{blue}{Epistasis to simulate order effects} {#two_five}

Epistasis assume that there is a dependence between genotypes. The effect of a mutation depends on the genetic background in which it happens [@Poelwijk2007]. Now, we will cope with dependencies between genes using epistasis. 

For that, we will use the same model described in \autoref{two_four}. As explained before, it is supposed to be a certain cancer progression restriction and therefore, the fitness values given to each different genotype is based in that criteria. 

```{r fig8, fig.align="center", fig.cap= "DAG showing epistasis between genes"}
## Fitness object defined using epistasias
cc_epi <- allFitnessEffects(epistasis = 
                              c("A: -B: -C" = 0.4,
                                "-A: B: -C" = -0.4, 
                                "-A: -B: C" = 0.3,
                                "A: B: -C" = 0.8,
                                "A : B: C" = 1.4,
                                "-A: B: C" = 0.1, 
                                "A : -B: C" = 0.5
                                ),
                            geneToModule =
                              c("A" = "APC",
                                "B" = "KRAS",
                                "C" = "TP53") 
                              )

## DAG (epistasis)
plot(cc_epi, expandModules = TRUE, autofit = TRUE)

## Genotypes derived from fitness defined with epistasia relationships
(cc_epi_geno <- evalAllGenotypes(cc_epi ))

```

```{r fig9, fig.align="center", fig.cap="Fitness landscape of model defined by epistasis"}

## Fitness landscape from this relationships
plotFitnessLandscape(cc_epi_geno, use_ggrepel = TRUE)

```
Using this approach, it is possible to visualize the DAG (see \autoref{fig:fig8}). In this case, there are discontinues yellow lines connecting each gene. This lines indicate a dependence between them. Fitness landscape is also plotted (see \autoref{fig:fig9}). 

With this model, we promote the clones of tumoral cells beginning with a first mutation in `APC`. Conversely, other clones not starting with that mutation (`KRAS` or `TP53`) have a lower fitness value. On the other hand, all genotypes end with the same fitness, but it is selective favored clones following the order defined in \autoref{two_four}. Hence, epistasis cannot be used to define order effects, but it can be used as a temporal solution.


# \textcolor{blue}{Pathway Linear Progression Model: Raphael \& Vanding, 2015} {#three}

The Pathway Linear Progression Model (PLPM) described in [@raphael_simultaneous_2015] introduces the idea that driver mutations target pathways. This is an important concept since different individuals have driver mutations in different genes that affect the same pathway [@raphael_simultaneous_2015]. Therefore, the order in which mutations arise are better described at the pathway level instead of a gene level [@raphael_simultaneous_2015]. 

Moreover, it has been previously described that there is a tendency of mutually exclusive mutations of genes participating in the same pathway and a tendency of co-occurring (overlapping) mutations of genes that affect different pathways [@yeang_combinatorial_2008]. Indeed, several cancer genes participate in multiple pathways and mutually exclusive events may allow for overlapping pathways [@Schill2020]. DISCUSS THIS LATER, MAYBE MODELS OF EVOLUTION IN ONCOSIMULR ALLOW FOR OVERLAPPING PATTERNS?

Here, we mapped the progression model from colorectal cancer data inferred by [@raphael_simultaneous_2015] (originally described in [@Wood2007]) into an evolutionary model, allowing deviations from the restriction imposed in the DAG. For this, we used a vector `s` to indicate the fitness effects when the restrictions are satisfied and a vector `sh` for deviations.

In [@raphael_simultaneous_2015], the authors analyzed eight genes: APC, EPHA3, EVC2, FBXW7,
KRAS, PIK3CA, TCF7L2, and TP53. In this model, APC mutations is an early event, followed by mutations in TP53 and PIK3CA (mutually exclusive). KRAS mutations appear after TP53/PIK3CA mutations. 

We used the `allFitnessEffects` function to  define the nodes and their relationships. Moreover, we used modules to represent mutually exclusive genes that affect the same pathway. Assigned fitness effects (`s`) values were higher for earlier mutations and lower for late mutation, since an earlier mutation in more prevalent in the clonal population than a later mutation, as explained in [@yeang_combinatorial_2008]. A single negative value was set for deviations from restrictions (`sh`) and a monotic relationship (`MN`) was used for relationships between nodes of the DAG since nodes have only one parent.   

\autoref{fig:fig10} shows the DAG inferred by [@raphael_simultaneous_2015] mapped to an evolutionary model that allows deviation from restrictions. Note that genes within a module are mutually exclusive and the restrictions goes top-down (i.e. from the root to the later mutation).

```{r, fig10, fig.align="center", fig.cap= "DAG from colorectal cancer dataset"}

## Define poset restrictions, mapping of genes to modules, and driver genes
CRC_W <- allFitnessEffects(data.frame(parent = c("Root", "A", "B", "C"),
                                      child = c("A", "B", "C", "D"),
                                      s = c(0.6, 0.4, 0.1, 0.05),
                                      sh = -0.5,
                                      typeDep = "MN"), 
                           geneToModule = c("Root" = "Root",
                                            "A" = "APC, EPHA3, TCF7L2",
                                            "B" = "EVC2, PIK3CA, TP53", 
                                            "C" = "KRAS",
                                            "D" = "FBXW7"),
                           drvNames = c("APC", "EPHA3", "TCF7L2", "EVC2", "PIK3CA", 
                                        "TP53", "KRAS", "FBXW7"))

# DAG representation
plot(CRC_W, expandModules = TRUE, autofit = TRUE, lwdf = 2)

```


The function `evalAllGenotypes` was used to map genotypes to fitness values. \autoref{fig:fig11} shows the fitness landscape inferred from the DAG of \autoref{fig:fig10}. As mentioned before, this fitness landscape with eight genes is difficult to visualize. Nevertheless, we can give a general description of the topology of the landscape. Note that there are multiple peaks and valleys, suggesting a high degree of ruggedness. Moreover, note that KRAS constitutes a local minima. This results confirms the order of restrictions imposed by the DAG. It is important to mention that some genotypes in the local maxima are composed of genes that belong to the same module. Such result may reflect driver genes that are mutually exclusive and participate in multiple pathways, previously described as overlapping pathways [@yeang_combinatorial_2008, @Schill2020]. 

```{r, fig11, fig.align="center", fig.cap= "Fitness landscape inferred from colorectal cancer DAG"}

## Map genotypes to fitness
CRC_F <- evalAllGenotypes(CRC_W, order = FALSE, addwt = TRUE)

## Plot of fitness landscape

plot(CRC_F)

```

## \textcolor{blue}{Simplified Model} {#three_one}

Given that our initial DAG contains eight genes, then number of possible genotypes is $2^{8} = 256$ which makes difficult to visualize the fitness landscape. For this, reason a smaller number of genes will be used to build a slightly different DAG to model other interesting scenarios (see \autoref{fig:fig12}). The idea is to represent mutual exclusivity with a `XOR` relationship (red edges). Also, note that the fitness value for mutual exclusive genes (APC and TP53) is almost the same.    

```{r, fig12, fig.align="center", fig.cap= "Simplified model from colorectal cancer DAG"}

## Simplified model
## Define poset restrictions, mapping of genes to modules, and driver genes
CRC_W2 <- allFitnessEffects(data.frame(parent = c(rep("Root", 2), "A", "B", "C"),
                                      child = c("A", "B", rep("C", 2), "D"),
                                      s = c(0.2, 0.1, rep(0.05, 2), 0.01),
                                      sh = -0.5,
                                      typeDep = c(rep("XMPN", 4), "MN")), 
                           geneToModule = c("Root" = "Root",
                                            "A" = "APC,",
                                            "B" = "TP53", 
                                            "C" = "KRAS",
                                            "D" = "FBXW7"),
                           drvNames = c("APC", "TP53", "KRAS", "FBXW7"))

# DAG representation
plot(CRC_W2, expandModules = TRUE, autofit = TRUE, lwdf = 2)

```


Since only four genes are used in the DAG, then the possible number of genotypes is $2^{4} = 16$, which are easier to interpret in a fitness landscape (see \autoref{fig:fig13}). Now, the genotypes with the highest fitness are the ones that fulfill the order of restrictions imposed by the DAG (e.g. APC, FBXW7, KRAS - APC, FBXW7, KRAS, TP53). On the other hand, genotypes that deviates from the imposed restrictions have the lowest fitness (e.g. KRAS - FBXW7 - APC, KRAS, TP53). However, specifying mutual exclusivity with `XOR` relationships cannot capture null effect or synthetic lethality between APC and TP53. Also, if an `AND` relationship is defined from the Root to APC and TP53, then there is no change in fitness values. 

```{r, fig13, fig.align="center", fig.cap= "Fitness landscape from simplified model"}

## Simplified Model
## Map genotypes to fitness
(CRC_F2 <- evalAllGenotypes(CRC_W2, order = FALSE, addwt = TRUE))

## Plot of fitness landscap

plot(CRC_F2, use_ggrepel = TRUE)


```

## \textcolor{blue}{SImulating Data from Simplified Model} {#three_three}

Fitness effects and restrictions defined in the DAG from \autoref{fig:fig12} from previous section was used to simulate clonal evolution. The same parameters from \autoref{two_three}, except `initSize` and  `finalTime`, were set in the `OncoSimulIndiv` function. \autoref{fig:fig14} shows the genotypes during that arises during clonal evolution. The genotype APC, TP53 fixates quickly in the clonal population. This result supports the fitness value for APC, FBXW7, KRAS depicted in \autoref{fig:fig13}, since that genotype is one of the local maxima. A more detailed order of genotype appearances and extinctions is shown in \autoref{fig:fig15}. Note that not all the 16 genotypes appear in the simulation because the best fitted genotype fixates rapidly in the population, leading to the extinction of some genotypes, whereas other cannot even appear. When simulation were executed with `onlyCancer = TRUE, cancer is never reached, although the fixated genotype is the global maxima of the fitness landscape (see \autoref{fig:fig16} and \autoref{fig:fig17}). 

```{r, fig14, fig.align="center", fig.cap= "Simulation of cancer progression for the simplified model. Genotypes are shown stacked"}

## Fix the seed for reproducibility
set.seed(87) 


CRC_W2_S <- oncoSimulIndiv(CRC_W2, ## A fitnessEffects object
  model = "McFL", ## Model used
  mu = 1e-4, ## Mutation rate
  sampleEvery = 0.02, ## How often the whole population is sampled
  keepEvery = 1,
  initSize = 2000, ## Initial population size
  finalTime = 2000,
  keepPhylog = TRUE, ## Allow to see parent-child relationships
  onlyCancer = FALSE)

## Plot of simulation for genotypes
plot(CRC_W2_S,
show = "genotypes",
type = "stacked")

```


```{r, fig15, fig.align="center", fig.cap= "Simulation of cancer progression for the simplified model. Genotypes are shown as lines."}

## Plot of simulation for genotypes
plot(CRC_W2_S,
show = "genotypes",
legend.ncols = 2,
xlim = c(0, 1500),
type = "line")

```

```{r, fig16, fig.align="center", fig.cap= "Simulation of cancer progression for the simplified model when onlyCancer = TRUE. Genotypes are shown stacked."}

## Fix the seed for reproducibility
set.seed(52) 


CRC_W2_S1 <- oncoSimulIndiv(CRC_W2, ## A fitnessEffects object
  model = "McFL", ## Model used
  mu = 1e-4, ## Mutation rate
  sampleEvery = 0.02, ## How often the whole population is sampled
  keepEvery = 1,
  initSize = 2000, ## Initial population size
  finalTime = 800,
  keepPhylog = TRUE, ## Allow to see parent-child relationships
  onlyCancer = TRUE,
  errorHitWallTime = FALSE, ## See results even if stopping conditions are not met
  errorHitMaxTries = FALSE)

## Time to reach cancer
(CRC_W2_S1$FinalTime)

## Plot of simulation for genotypes
plot(CRC_W2_S1,
show = "genotypes",
legend.ncols = 2,
xlim = c(0, 500),
type = "stacked")

```

```{r, fig17, fig.align="center", fig.cap= "Simulation of cancer progression for the simplified model when onlyCancer = TRUE. Genotypes are shown as lines."}

## Plot of simulation for genotypes
plot(CRC_W2_S1,
show = "genotypes",
legend.ncols = 1,
xlim = c(0, 300),
type = "line")

```


\autoref{fig:fig18} and \autoref{fig:fig19} shows the genealogical relationships of clones that appeared during the simulations. The number of the arrows represent the times that each clone appeared. When simulation are set to reach cancer the clones that have a genotype belonging to the two local optima appear(\autoref{fig:fig19}). Whereas if simulation are executed without the cancer parameter, the most represented clone is the one that has the best fitted genotype (see \autoref{fig:fig18}).

```{r, fig18, fig.align="center", fig.cap= "Genealogical relationships of clones."}

## Plot of genealogical relationships
plotClonePhylog(CRC_W2_S, N = 0, keepEvents = TRUE)

```


```{r, fig19, fig.align="center", fig.cap= "Genealogical relationships of clones when onlyCancer = TRUE."}

## Plot of genealogical relationships
plotClonePhylog(CRC_W2_S1, N = 0, keepEvents = TRUE)

```

## \textcolor{blue}{Synthetic Lethality} {#three_four}

Synthetic lethality is a special type of epistasis. Therefore, we used the epistasis module inside `allFitnessEffects` to define an epistatic interaction between TP53 and AP (see \autoref{fig:fig20}) and restriction imposed by the DAG (i.e. `XOR` relationships). The fitness values were assigned such that an scenario where synthetic lethality via pairwise interaction occurs. 

The fitness landscape shows that the genotype for which the synthetic lethality was specified has a lower fitness value as expected, although it is not a local minima. Similarly to fitness landscape in \autoref{fig:fig13}, the local maxima is composed by the genotyopes that satisfy both epistatic interactions and restrictions imposed. Whereas, local minima is composed by genotypes that contain genes with synthetic lethality and other genes that have top-down dependencies (see \autoref{fig:fig21}).  

```{r, fig20, fig.align="center", fig.cap= "DAG with synthetic lethality."}

## Simplified model
## Define poset restrictions, mapping of genes to modules, and driver genes
CRC_W3 <- allFitnessEffects(data.frame(parent = c(rep("Root", 2), "A", "B", "C"),
                                      child = c("A", "B", rep("C", 2), "D"),
                                      s = c(0.2, 0.1, rep(0.05, 2), 0.01),
                                      sh = -0.5,
                                      typeDep = c(rep("XMPN", 4), "MN")), 
                            epistasis = c("-A : B" = 0.1,
                                          "-B : A" = 0.2,
                                          "A:B" = -0.5),
                           geneToModule = c("Root" = "Root",
                                            "A" = "APC",
                                            "B" = "TP53", 
                                            "C" = "KRAS",
                                            "D" = "FBXW7"),
                           drvNames = c("APC", "TP53", "KRAS", "FBXW7"))

# DAG representation
plot(CRC_W3, expandModules = TRUE, autofit = TRUE, lwdf = 2)

```

```{r, fig21, fig.align="center", fig.cap= "Fitness landscape inferred from simplified DAG with synthetic lethality."}

## Map genotypes to fitness
CRC_F1 <- evalAllGenotypes(CRC_W3, order = FALSE, addwt = TRUE)

(CRC_F1)

## Plot of fitness landscape

plot(CRC_F1, use_ggrepel = TRUE)

```


In order to simulate synthetic lethality via three-way interaction, we set fitness values that reflect slightly deleterious effect (if two genes appear) or a highly deleterious effect (if three genes appear). \autoref{fig:fig22} shows the DAG derived for the three-way interaction between APC, TP53, and KRAS. The inferred fitness landscape shows that the global minima is composed by the genotype that carries the synthetic lethality. Whereas, local maxima is composed by genotypes that follow the restrictions imposed in the DAG. Also, note that the global maxima is APC. This is not surprising given that APC is an earlier mutation and has the highest fitness values compared to other genes/genotypes (see \autoref{fig:fig23}).    

```{r, fig22, fig.align="center", fig.cap= "DAG with synthetic lethality (three-way interaction)."}

## Simplified model
## Define poset restrictions, mapping of genes to modules, and driver genes
CRC_W4 <- allFitnessEffects(data.frame(parent = c(rep("Root", 2), "A", "B", "C"),
                                      child = c("A", "B", rep("C", 2), "D"),
                                      s = c(0.2, 0.1, rep(0.05, 2), 0.01),
                                      sh = -0.5,
                                      typeDep = c(rep("XMPN", 4), "MN")), 
                            epistasis = c("A : -B : -C" = 0.2,
                                          "-A : B : -C" = 0.1,
                                          "-A : -B : C" = 0.05,
                                          "A : B : -C" = 0.01,
                                          "-A : B : C" = 0.02,
                                          "-B : A : C" = 0.02,
                                          "A : B : C" = -0.5),
                           geneToModule = c("Root" = "Root",
                                            "A" = "APC",
                                            "B" = "TP53", 
                                            "C" = "KRAS",
                                            "D" = "FBXW7"),
                           drvNames = c("APC", "TP53", "KRAS", "FBXW7"))

# DAG representation
plot(CRC_W4, expandModules = TRUE, autofit = TRUE, lwdf = 2)

```


```{r, fig23, fig.align="center", fig.cap= "Fitness landscape inferred from simplified DAG with synthetic lethality (three-way interaction)."}

## Map genotypes to fitness
CRC_F2 <- evalAllGenotypes(CRC_W4, order = FALSE, addwt = TRUE)

(CRC_F2)

## Plot of fitness landscape

plot(CRC_F2, use_ggrepel = TRUE)

```

## \textcolor{blue}{Synthetic Viability} {#three_four}

Synthetic viability is specified for genotype APC, TP53 (see \autoref{fig:fig24}). Here the genotypes composed only by APC or TP53 are deleterious. \autoref{fig:fig25} shows the fitness landscape for synthetic viability via pairwise interaction. Note that the global maxima is composed by the genotype that contains all genes. On the other hand, local minima is composed by genotypes that contains one gene that has a deleterious effect. Note that despite the lower fitness value of genotype FBXW7, KRAS, it conforms a local maxima, although the restrictions imposed in the DAG are not completely satisfied. Moreover, in this fitness landscape, the global maxima may not be reached because the mutational paths required lead to a region composed of multiplpe valleys. It important to mention that order of effects could provide a more realistic fitness landscape. For example, a possible path that leads to the global maxima requires a mutation in KRAS before a mutation in FBXW7.  


```{r, fig24, fig.align="center", fig.cap= "DAG with synthetic viability (pairwise interaction)."}

## Simplified model
## SM because synthetic viability requires both parent nodes.
## Define poset restrictions, mapping of genes to modules, and driver genes
CRC_W5 <- allFitnessEffects(data.frame(parent = c(rep("Root", 2), "A", "B", "C"),
                                      child = c("A", "B", rep("C", 2), "D"),
                                      s = c(0.2, 0.1, rep(0.05, 2), 0.01),
                                      sh = -0.5,
                                      typeDep = c(rep("MN", 5))), 
                            epistasis = c("-A : B" = -0.2,
                                          "-B : A" = -0.3,
                                          "A:B" = 0.5),
                           geneToModule = c("Root" = "Root",
                                            "A" = "APC",
                                            "B" = "TP53", 
                                            "C" = "KRAS",
                                            "D" = "FBXW7"),
                           drvNames = c("APC", "TP53", "KRAS", "FBXW7"))

# DAG representation
plot(CRC_W5, expandModules = TRUE, autofit = TRUE, lwdf = 2)

```

```{r, fig25, fig.align="center", fig.cap= "Fitness landscape inferred from simplified DAG with synthetic viability."}

## Map genotypes to fitness
CRC_F3 <- evalAllGenotypes(CRC_W5, order = FALSE, addwt = TRUE)
(CRC_F3)

## Plot of fitness landscape

plot(CRC_F3, use_ggrepel = TRUE)

```


\autoref{fig:fig26} shows synthetic viability with a three-way interaction between APC, TP53, and KRAS. For this, we specified highly deleterious effects if APC, TP53, or KRAS appear independently. Whereas, slightly deleterious effects were set if two of those genes appear in a genotype. The fitness landscape for this scenario (see \autoref{fig:fig27}) shows the order of restrictions and epistatic interactions set lead to the global maxima composed by the genotype APC, TP53, KRAS, FBXW7. This result support the idea that DAGs are better suited to represent sign epistasis [@Diaz-Uriarte2018]. Nevertheless, as mentioned above, including order of effects can give a more realistic fitness values associated to genotypes. 

In this work, we have represented synthetic lethality via pairwise and three-way interactions. However, this can be achieved if the DAG is composed by individual genes instead of modules because modules does not allow to define epistatic relationships between genes of the same module. This is important because genes of the same module can participate in the same pathway, as discussed in [@Schill2020, @raphael_simultaneous_2015]. 

```{r, fig26, fig.align="center", fig.cap= "DAG with synthetic viability (three-way interaction)."}

## Simplified model
## SM because synthetic viability requires both parent nodes. 
## Define poset restrictions, mapping of genes to modules, and driver genes
CRC_W6 <- allFitnessEffects(data.frame(parent = c(rep("Root", 2), "A", "B", "C"),
                                      child = c("A", "B", rep("C", 2), "D"),
                                      s = c(0.2, 0.1, rep(0.05, 2), 0.01),
                                      sh = -0.5,
                                      typeDep = c(rep("MN", 5))), 
                            epistasis = c("A : -B : -C" = -0.2,
                                          "-A : B : -C" = -0.2,
                                          "-A : -B : C" = -0.3,
                                          "A : B : -C" = -0.05,
                                          "-A : B : C" = -0.01,
                                          "A : -B : C" = -0.01,
                                          "A : B : C" = 0.5),
                           geneToModule = c("Root" = "Root",
                                            "A" = "APC",
                                            "B" = "TP53", 
                                            "C" = "KRAS",
                                            "D" = "FBXW7"),
                           drvNames = c("APC", "TP53", "KRAS", "FBXW7"))

# DAG representation
plot(CRC_W6, expandModules = TRUE, autofit = TRUE, lwdf = 2)

```

```{r, fig27, fig.align="center", fig.cap= "Fitness landscape inferred from simplified DAG with synthetic viability (three-way interaction)."}

## Map genotypes to fitness
CRC_F4 <- evalAllGenotypes(CRC_W6, order = FALSE, addwt = TRUE)
(CRC_F4)

## Plot of fitness landscape

plot(CRC_F4, use_ggrepel = TRUE)

```

# \textcolor{blue}{A probabilistic model of mutually exclusive linearly ordered driver pathways} {#four}

Mohaghegh Neyshabouri et al. @Neyshabouri2020 propose a probabilistic model of mutually exclusive linearly ordered driver pathways and analyze one large dataset of colorectal adenocarcinoma (COADREAD) from IntOGen-mutations database. Their model assumes driver genes are over-represented among those mutated across a large tumor collection and, thus, they can be identified in terms of frequency. Also, those participating of the same pathway are mutated in a mutually exclusive manner because more than one mutation in a pathway does not give any selective advantage to the clone.

Like with previous generative models, we map the COADREAD generative model to an actual evolutionary model using different **OncoSimulR** functionalities. This time, we extent what authors model using the frequency-dependent fitness specification, to illustrate how differently fitness landscapes evolve even though they are built from exact CPMs when we consider this additional evolutionary event 

Figure 7.C from @Neyshabouri2020 shows the CPM inferred from the COADREAD dataset, consisting of seven modules with between 1 to 4 genes each. The model clearly reconstructs the well-known initiator events in colorectal cancer, including mutations in *APC*, *TP53* and *KRAS* @Gerstung2011. Using the DAG of restrictions as starting point[^1], the evolutionary model is created specifying same genotype fitness for all modules as authors do not state any differences in fitness for when the restrictions in the DAG are satisfied (`s`). However, based on the confidence parameter used by the authors to assess the reliability of modeled restrictions, different fitness are set when the DAG of restrictions is not satisfied (`sh`) (\autoref{table2}). Since this method reconstructs linear models (*i.e.* oncogenic trees), there is no need to specify any particular type of dependency between modules (`typeDep`), so we set it to monotonic (MN) as it is a mandatory argument for `allFitnessEffects` function. \autoref{fig:fig28} shows the DAG of restrictions created with `allFitnessEffects`, recapitulating the poset inferred in @Neyshabouri2020.

[^1]: Several genes were removed from the original set in order to get a clear fitness landscape

Table: confidence parameter for each module transition \label{table2}

|                   Module                  |   Confidence parameter (%)    |
| :---------------------------------------: | :---------------------------: |
|                   APC                   |             100               |
|                                           |                               |
|                   TP53                  |             100               |
|                                           |                               |
|                   KRAS                  |             100               |
|                                           |                               |
|         PIK3CA, NRAS, LRP1B         |             100               |
|                                           |                               |
|     FBXW7, TCF7L2, FAT4, ARID1A   |             87.7              |
|                                           |                               |
| ATM, SMAD2, ERBB3, MTOR, CTNNB1 |             86.9              |
|                                           |                               |
|               SOX9, SMAD4             |             66.7              |

```{r fig28, fig.align="center", fig.cap= "DAG of restrictions for the COADREAD dataset"}

## Restriction table, including DAG of restrictions specifications and associated fitness
COADREAD_rT <- data.frame(parent = c("Root", "A", "B", "C", "D", "E", "F"), # Parent nodes
                             child = c("A", "B", "C", "D", "E", "F", "G"), # Child nodes
                             s = 0.5, 
                             sh = c(rep(-1, 4), rep(-.5, 2), -.2),
                             typeDep = "MN")


## Create fitness specifications from DAG of restrictions considering modules 
COADREAD_fitness <- allFitnessEffects(COADREAD_rT, 
                                         geneToModule = c( "Root" = "Root",
                                                           "A" = "APC",
                                                           "B" = "TP53",
                                                           "C" = "KRAS",
                                                           "D" = "PIK3CA, NRAS",
                                                           "E" = "FBXW7, ARID1A",
                                                           "F" = "ATM, SMAD2",
                                                           "G" = "SOX9, SMAD4")) # Modules


## DAG of restrictions representation
plot(COADREAD_fitness, expandModules = TRUE, autofit = TRUE)
```
```{r fig29, fig.align="center", fig.cap= "Fitness landscape corresponding with the DAG of restrictions for the COADREAD dataset"}
# Evaluation of all possible genotypes fitness under the previous fitness specifications
COADREAD_FL <- evalAllGenotypes(COADREAD_fitness, max = 131072)

# Fitness landscape representation
plotFitnessLandscape(COADREAD_FL) 
```
\autoref{fig:fig29} plots all possible genotypes in a very busy fitness landscape. Although it is not possible to visualize genotypes clearly, we can see an exponential trend towards local maxima, corresponding to the clones carrying seven driver-genes genotypes. Stands out how very different genotypes behave the same due to the mutual exclusivity effect, which is probably simplifying much more complex interactions among genes, specially the effect in fitness different genes from the same module could have if we didn't account for the mutual exclusivity phenomenon. In next section we discuss genotypes distribution with a simplified model. 

## \textcolor{blue}{Simplified cancer progression model} {#four_one}

As we did with previous models, for illustrating purposes we designed a simplified version of the CPM by @Neyshabouri2020 to assess its reliability when considering other evolutionary scenarios. This time, we take the first four modules of the complete DAG (mantaining same fitness parameters), being the fourth the only one carrying two genes and, thus, affected by mutual exclusivity circumstance (\autoref{fig:fig30}. 

The fitness landscape shown in \autoref{fig:fig31} includes some of the possible genotypes. Clearer than in \autoref{fig:fig29}, we see fitness increases exponentially following the restrictions established in the DAG. Local minima (red squares) correspond to genotypes violating these constraints, and gain changes (green lines) are consistent with satisfied restrictions. However, one of the local maxima genotypes (green squares) corresponds to a genotype violating mutual exclusivity (PIK3CA and NRAS appear together). This is explained because the Modules functionality in OncoSimulR sets fitness to cero when genes from the same module mutate at a time as there is a null effect because both genes participate from the same pathway. Because there is not a decrease in the fitness of the genotype (*i.ei* no deleterious effect), it can still be a local maxima. Yet, it is unlikely a genotype needed of an extra mutation to compete with the other maximum-fitness genotypes will survive long during cancer progression to fixate. DE TODAS FORMAS, AQUÍ IRÍA LO QUE SEA QUE NOS DIGA RAMÓN. 

```{r fig30, fig.align="center", fig.cap= "Simplified DAG of restrictions for the COADREAD dataset"}

## Restriction table, including five-modules DAG of restrictions specifications and associated fitness
COADREAD_rT_5d <- data.frame(parent = c("Root", "A", "B", "C"), # Parent nodes
                             child = c("A", "B", "C", "D"), # Child nodes
                             s = 0.5, 
                             sh = c(rep(-1, 4)),
                             typeDep = "MN")

## Create fitness specifications from simplified DAG of restrictions
COADREAD_fitness_5d <- allFitnessEffects(COADREAD_rT_5d,
                                         geneToModule = c( "Root" = "Root",
                                                           "A" = "APC",
                                                           "B" = "TP53",
                                                           "C" = "KRAS",
                                                           "D" = "PIK3CA, NRAS"),
                                         drvNames = c("APC", "TP53", "KRAS",
                                                      "PIK3CA", "NRAS"))

## Simplified DAG of restrictions representation
plot(COADREAD_fitness_5d, expandModules = TRUE, autofit = TRUE)
```
```{r fig31, fig.align="center", fig.cap= "Fitness landscape corresponding with the simplified DAG of restrictions for the COADREAD dataset"}
# Evaluation of all possible genotypes fitness under the previous fitness specifications
COADREAD_FL_5d <- evalAllGenotypes(COADREAD_fitness_5d)

# Fitness landscape representation
plotFitnessLandscape(COADREAD_FL_5d, use_ggrepel = TRUE) 
plotFitnessLandscape(COADREAD_FL_5d) 

```
The previous assumption can be better addressed simulating tumor progression under the proposed evolutionary model. Accordingly, we would expect clones to evolve following the optimal pathway of the fitness landscape, that is satisfying the DAG of restrictions. We use `oncoSimulPop` function to simulate tumor progression in ten different individuals and test whether the genotype `APC, TP53, KRAS, PIK3CA, NRAS` would ever fixate even at a very small frequency. Parameters are adjusted appropriately so that simulation stops when `APC, TP53, KRAS, PIK3CA, NRAS` genotype is reached at a frequency of 0.1 at least (`onlyCancer = TRUE, fixation = c(c("_, APC, TP53, KRAS, PIK3CA, NRAS"), fixation_tolerance = 0.9)`). 

\autoref{fig:fig32} shows a representative stacked plot of the genotype abundance over time. Surprisingly, `APC, TP53, KRAS, PIK3CA, NRAS` genotype appeared in the ten simulations significantly (Data not shown), able to coexist with clones of the same fitness. Often, it arises from `APC, TP53, KRAS, PIK3CA` or `APC, TP53, KRAS, NRAS` clones, which appear together even less frequently, probably because both descend from the same ancestor and have much higher fitness, leading to its extinction once one of the two emerges. \autoref{fig:fig33} supports this idea as it clearly plots the behavior of each clone separately, displaying abrupt extinctions the moment the following clone with higher fitness appears. Retrieving the phylogeny of the clones (representative graph of one simulation in \autoref{fig:fig34}), we confirm there is a tendency of the clones to mutate until the complete genotype `APC, TP53, KRAS, PIK3CA, NRAS`, either following the linear mutual exclusivity path or from a more diverse landscape. 

```{r fig32, fig.align="center", fig.cap= "One of the 10 simulations of cancer progression using the four-modules fitness landscape until APC, TP53, KRAS, PIK3CA, NRAS genotype arises (stacked plot)"}
set.seed(125)

# Simulate cancer progression in 10 individuals until APC, TP53, KRAS, PIK3CA, NRAS genotype fixates
COADREAD_Simul_5d <- oncoSimulPop(10, COADREAD_fitness_5d, 
                           model = "McFL", ## Model used
                           mu = 1e-4, ## Mutation rate
                           sampleEvery = 0.02, ## How often the whole population is sampled
                           keepEvery = 1,
                           initSize = 200, ## Initial population size
                           finalTime = 2000,
                           keepPhylog = TRUE, ## Allow to see parent-child relationships
                           onlyCancer = TRUE,
                           detectionSize = NA,
                           fixation = c(c("_, APC, TP53, KRAS, PIK3CA, NRAS"),
                                        fixation_tolerance = 0.7),
                           detectionDrivers = NA,
                           detectionProb = NA,
                           max.num.tries = 500,
                           max.wall.time = 20,
                           errorHitMaxTries = TRUE)

## Plot of simulation for genotypes
plot(COADREAD_Simul_5d[[3]],
show = "genotypes",
type = "stacked",
ylim = c(0, 35000))
```

```{r fig33, fig.align="center", fig.cap= "One of the 10 simulations of cancer progression using the four-modules fitness landscape until APC, TP53, KRAS, PIK3CA, NRAS genotype arises (line plot)"}
plot(COADREAD_Simul_5d[[5]], 
     show = "genotypes",
     type = "line",
     ylim = c(1, 100000000))
```
```{r fig34, fig.align="center", fig.cap="Parent-child relationship derived from one of the 10 simulations"}
## Parent-child relationship derived from simulation
plotClonePhylog(COADREAD_Simul_5d[[3]],
                N = 0, ## Specify clones that exist
                keepEvents = TRUE ## Arrows showing how many times each clones appeared
)
```


## \textcolor{blue}{Frequency-dependent fitness} {#four_two}

As previously introduced, clones that coexist in a tumor can influence the fitness of each other in a frequency-dependent manner when a mutation produces a phenotype able to modulate the tumor microenvironment. OncoSimulR incorporates the `frecuencyDependentFitness` specification to allow modelling interactions among clones during tumor progression. In the simulations ran with the simplified model, we significantly observed `APC, TP53, KRAS, PIK3CA, NRAS` genotype in the final stage of the simulation, an unexpected event considering `PIK3CA` and `NRAS` are mutually exclusive. This might denote there is an additional evolutionary event directing the very infrequent coexistence of these two genes in tumor samples (@Neyshabouri2020). Here, we propose `APC, TP53, KRAS, PIK3CA, NRAS` genotype fitness could depend on the frequency of `APC, TP53, KRAS, PIK3CA` and `APC, TP53, KRAS, NRAS` genotypes in the context of a competitive relationship among clones for niche nutrients. `APC, TP53, KRAS, PIK3CA, NRAS` clones would be more energetically demanding and thus, coexistence with other clones would be detrimental (considering the three clones have same fitness).

To use the `frecuencyDependentFitness` functionality, it is necessary to set to TRUE de `frequencyDependentFitness` parameter in the `allFitnessEffects` function, as well as providing a "mapping of genotypes to fitness" data frame. Fitness values are taken from the fitness specifications previously used (\autoref{four_one_two}). To evaluate genotypes with the `evalAllGenotypes` function is mandatory the parameter `spPopSizes` to build a fitness landscape in accordance with the size of the different clones. Here, we define a scenario in which all the clones except for the four-genes and five-genes clones are almost extincted. In this context, we already see in the fitness landscape (\autoref{fig:fig35}) an expected fitness decreased in `APC, TP53, KRAS, PIK3CA, NRAS` genotype, which is not a local maxima anymore.  



```{r fig35, fig.align="center", fig.cap="Fitness landscape corresponding with the simplified DAG for the COADREAD dataset accounting for frequency-dependent fitness"}

## Mapping of genotypes to frequency-dependent fitness
# Not explicitly mapped genotypes are assigned a fitness of cero
COADREAD_gen_freqdep <- data.frame(Genotype = c("WT", "APC","APC, TP53",
                                                "APC, TP53, KRAS",
                                                "APC, TP53, KRAS, PIK3CA",
                                                "APC, TP53, KRAS, NRAS",
                                                "APC, TP53, KRAS, PIK3CA, NRAS"),
                            Fitness = c("1", "1.5", "2.25", "3.375", "5.0625", "5.0625",
                                        "5.0625 - ((f_APC_TP53_KRAS_PIK3CA + f_APC_TP53_KRAS_NRAS))/2"),
                            stringsAsFactors = FALSE)

## Fitness specifications

COADREAD_fitness_freqdep <- allFitnessEffects(genotFitness = COADREAD_gen_freqdep, 
                                                   frequencyDependentFitness = TRUE,
                                                   frequencyType = "rel")

## Evaluate all genotypes considering population sizes of the clones
COADREAD_FL_freqdep <- evalAllGenotypes(COADREAD_fitness_freqdep,
                                 spPopSizes = c("WT" = 5, "APC" = 5, "APC, TP53" = 5,
                                                "APC, TP53, KRAS" = 10,
                                                "APC, TP53, KRAS, PIK3CA" = 50,
                                                "APC, TP53, KRAS, NRAS" = 50,
                                                "APC, TP53, KRAS, PIK3CA, NRAS" = 50))
                                                   

# Fitness landscape representation
plotFitnessLandscape(COADREAD_FL_freqdep) 
```
Next, we try to run the same simulation we did before with these new fitness specifications. After modifying the `fixation_tolerance` parameter so that we could detect the `APC, TP53, KRAS, PIK3CA, NRAS` genotype rapidly, it never arises. This can be explained by its slightly lower fitness compare to its ancestors, which by the moment `PIK3CA` or `NRAS` mutate to generate the fivefold-mutated genotype have a very high frequency, slowing down `APC, TP53, KRAS, PIK3CA, NRAS` clone growth up to extinction. Running an additional short simulation, in which we just set `onlyCancer` as FALSE (so as for the simulation to run until `finalTime`), in \autoref{fig:fig36} we don't see the presence of the fivefold-mutated clone, yet \autoref{fig:fig37} shows an oscillating pattern of growth for this clone. Also, the phylogeny recorded for each simulation shows the appearance of `APC, TP53, KRAS, PIK3CA, NRAS` genotype (\autoref{fig:fig38}. Either way, its frequence is so low that it cannot trigger the `onlyCancer` condition to end the first simulation. Although we cannot assure whether the phenomenon of frequency-dependent fitness could be influencing on the lack of coexistence of `PIK3CA` and `NRAS` in real colorectal cancer samples (@Neyshabouri2020), these findings support there are probably additional evolutionary events leading genotype frequency apart from mutual exclusivity. 
```{r}
# Simulate cancer progression in 10 individuals until APC, TP53, KRAS, PIK3CA, NRAS genotype fixates

set.seed(125)
COADREAD_Simul_freqdep <- oncoSimulIndiv(COADREAD_fitness_freqdep, 
                           model = "McFL", ## Model used
                           mu = 1e-4, ## Mutation rate
                           sampleEvery = 0.02, ## How often the whole population is sampled
                           keepEvery = 1,
                           initSize = 200, ## Initial population size
                           finalTime = 2000,
                           keepPhylog = TRUE, ## Allow to see parent-child relationships
                           onlyCancer = TRUE,
                           detectionSize = NA,
                           fixation = c(c("_, APC, TP53, KRAS, PIK3CA, NRAS"),
                                        fixation_tolerance = 0.99),
                           detectionDrivers = NA,
                           detectionProb = NA)
```
```{r fig36, fig.align="center", fig.cap= "One of the 10 simulations of cancer progression using the frequency-dependent fitness model for the COADREAD dataset (stacked plot)"}
# Simulate cancer progression in 10 individuals for a final time of 300 time units

set.seed(125)
COADREAD_Simul_freqdep <- oncoSimulPop(10, COADREAD_fitness_freqdep, 
                           model = "McFL", ## Model used
                           mu = 1e-4, ## Mutation rate
                           sampleEvery = 0.02, ## How often the whole population is sampled
                           keepEvery = 1,
                           initSize = 200, ## Initial population size
                           finalTime = 300,
                           keepPhylog = TRUE, ## Allow to see parent-child relationships
                           onlyCancer = FALSE,
                           detectionSize = NA,
                           fixation = NA,
                           detectionDrivers = NA,
                           detectionProb = NA)

## Plot of simulation for genotypes
plot(COADREAD_Simul_freqdep[[9]],
show = "genotypes",
type = "stacked",
ylim = c(0, 35000))

```
```{r fig37, fig.align="center", fig.cap= "One of the 10 simulations of cancer progression using the frequency-dependent fitness model for the COADREAD dataset (line plot)"}
plot(COADREAD_Simul_freqdep[[9]], 
     show = "genotypes",
     type = "line",
     ylim = c(1, 100000000))
```
```{r fig38, fig.align="center", fig.cap="Parent-child relationship derived from one of the 10 simulations"}
## Parent-child relationship derived from simulation
plotClonePhylog(COADREAD_Simul_freqdep[[9]],
                N = 0, ## Specify clones that exist
                keepEvents = TRUE ## Arrows showing how many times each clones appeared
)
```


# \textcolor{blue}{References} {#ref}

<div id="refs"></div>

